Foregrounds in Fourier Space for Redshifted 21~cm Observatories

### 2015-02-04 ms.tex
Make sec. 2 compatible with alternate outline
from 379e27b:
============

In ms.tex,

(*) Make small changes to end of section 2
(*) Make small changes to caption of figure 2
(*) Remove reference to Fourier space figure as being specific to MWA
(*) Now sections 2, 3, and 4 are more or less compatible with the
alternate outline

Equation~\ref{eqn:obsvis} can be {\bf equivalently re-expressed} as:
\begin{align}\label{eqn:vis}
  V_{uf} &\equiv V_{uf}(\boldsymbol{u},f) \nonumber\\
  &= \iint\limits_\textrm{sky} A(\hat{\boldsymbol{s}},f)\,I(\hat{\boldsymbol{s}},f)\,W_f\,e^{-i2\pi\boldsymbol{u}\cdot\hat{\boldsymbol{s}}}\,\dif\Omega,
\end{align}
where, $\hat{\boldsymbol{s}}$ is measured with reference to a location on the sky referred to as the {\it phase center}, and $\boldsymbol{u}\equiv (u,v,w)$ denotes the spatial frequency vector. $w$ is aligned parallel to the direction of the phase center, while $u$ and $v$ lie on the transverse plane perpendicular to it. For measurements that are constrained to lie on this plane, we can choose $w=0$ without loss of generality and $\boldsymbol{u}$ effectively reduces to $\boldsymbol{u}\equiv (u,v)$, a two--dimensional vector. Then, $\boldsymbol{u}$ is directly related to the transverse spatial wavenumber mode as: 
\begin{align}\label{eqn:k-perp}
  \boldsymbol{k}_\perp &\equiv \frac{2\pi\boldsymbol{u}}{D}, 
\end{align}
where, $D\equiv D(z)$ is the transverse comoving distance at redshift $z$. % Hence, $V_{uf}=V_{uf}(\boldsymbol{b}/\lambda,f)=V_{bf}(\lambda\boldsymbol{u},f)$. It is important to note that for a given antenna pair, $\boldsymbol{b}$ is independent of frequency while $\boldsymbol{u}$ is not.

@@

Without loss of generality, the phase center can be assumed to be the zenith relative to the local tangent plane. Then $\boldsymbol{u}$ lies on this plane for measurements constrained to be on it. If the array of antennas are also coplanar lying on the local tangent plane, then $\boldsymbol{u}=\boldsymbol{b}/\lambda$. Under such circumstances, equations~\ref{eqn:obsvis} and \ref{eqn:delay} closely resemble equations~\ref{eqn:vis} and \ref{eqn:los-transform} respectively. However, they are not quite identical to each other. The difference is because $\boldsymbol{b}$ is independent of frequency while $\boldsymbol{u}$ is not. \citet{par12b} and \citet{liu14a} have discussed the mathematical correspondence between the two.

@@

{\bf We give a brief overview of some parameters of Fourier space which is generic to all experiments that use a similar approach}. Figure~\ref{fig:fourier-space} illustrates the Fourier space, {\bf specific to the MWA}, in which the delay (and power) spectra of H{\sc i} from the EoR are presented in this paper. $|\boldsymbol{b}|$ and $k_\perp$, denoting spatial scales in the transverse direction {\bf (tangent plane to the celestial sphere)}, form the $x$--axis. $\tau\approx\eta\propto k_\parallel$, denoting spatial scales along line--of--sight form the $y$--axis. Foreground emission maps to a wedge--shaped region in Fourier space, hereafter referred to as the {\it foreground wedge} \citep{dat10}, whose boundaries are determined by the antenna spacings and the light travel times across them. These boundaries, called horizon delay limits \citep{ved12,par12b}, are shown by solid lines. The spectral transfer function of the instrument convolves the {\it foreground wedge} and stretches it further (unshaded narrow strips bounded by solid and dashed lines) along $\tau$--axis \citep{par12b,thy13}. The width of this narrow strip is inversely proportional to the operating bandwidth. The region of Fourier space excluding the {\it foreground wedge} and the narrow strips is the so--called {\it EoR window}, shown in light and medium shades. However, the instrument used in our study, namely the MWA, has a passband constructed using coarse channels with a characteristic shape giving rise to periodic gratings thereby resulting in repetitions of the {\it foreground wedge} at multiples of 0.78~$\mu$s.

In the context of EoR studies in Fourier space, the H{\sc i} power spectrum from the EoR is expected to decrease rapidly with $|\boldsymbol{k}|$. Hence, the brightest EoR signal will be observed on the shortest baselines and smallest delays. Thus the region of interest for EoR studies is slightly beyond the horizon delay limits (dashed lines) and inside the first grating on either side ($|\tau| \lesssim 0.78\,\mu$s, dot--dashed lines) caused by the instrument's spectral transfer function. We call this region with medium shade as the {\it EoR window} in this paper. The MWA EoR collaboration actively uses this as a test region in Fourier space for diagnostics on data quality.

\begin{figure}[htb]
\centering
% \includegraphics[width=\linewidth]{figures/v1_0/fourier_space_185.0_MHz_30.7_MHz.eps}
% \includegraphics[width=\linewidth]{f1.eps}
% \includegraphics[width=\linewidth]{fig2.eps}
\includegraphics[width=\linewidth]{figure2.eps}
\caption{Fourier space in which delay (and power) spectra of EoR H{\sc i} signals are presented. The $x$--axis is denoted by $|\boldsymbol{b}|$ (antenna spacing) or $k_\perp$ (transverse wavenumber). The $y$--axis denoted by $\tau$ (delay) or $k_\parallel$ (line--of--sight wavenumber). Here, $k_\perp$ and $k_\parallel$ are obtained for a frequency of 185~MHz. The dark shaded region is referred to as the {\it foreground wedge} where smooth spectrum foregrounds reside. Its boundaries (solid lines), given by light travel time for corresponding antenna spacings, are referred to as horizon delay limits. Regions excluding the {\it wedge} and narrow extensions of the {\it wedge} (white unshaded strips) caused by convolution with the {\bf instrument's spectral transfer function} are expected to be relatively free of foreground contamination. Due to an undesirable grating pattern (dot--dashed lines) of the instrument used in this study, namely the MWA, we conservatively identify a restricted region of high EoR sensitivity (medium shade) and refer to it as the {\it EoR window} in this paper. \label{fig:fourier-space}}
\end{figure}

In our present study, we use a latitude of -26.701\arcdeg and an antenna layout identical to that of the MWA \citep{bea12}. The array is arranged as a centrally condensed core of $\sim$~300~m --- there are many spacings in the range 5--50~m --- and a radial density that falls off as the inverse of the radius, with the longest baselines at 3~km. Here we focus on antenna spacings $|\boldsymbol{b}| \le 200$~m (spatial scales relevant to reionization). For geometrical intuition, we restrict the orientation ($\theta_b$, measured anti-clockwise from East) of all baselines to lie in the range $-67.5\arcdeg \leq \theta_b < 112.5\arcdeg$. Baselines oriented in the other half--plane measure conjugate visibilities with delays of equal magnitude but of opposite sign and hence are ignored in our analysis. {\bf Note that visibilities from different baselines have not been averaged together in this analysis}.

@@

\begin{figure}[htb]
\centering
\includegraphics[width=\linewidth]{figure3.eps}
\caption{Delay power spectra of a uniform sky brightness distribution measured by antenna pairs with a uniform power pattern across the visible hemisphere. The non--zero response of the interferometer array to a uniform brightness distribution and the prominent {\it edge brightening} close to the horizon delay limits are due to wide--field effects. These effects are prevalent on all antenna spacings and are generic to all instruments used in wide--field measurements. \label{fig:usm}}
\end{figure}

We interpret this is due to equal--sized delay bins subtending larger solid angles near the horizon thereby containing larger integrated emission. An alternate, but equivalent, interpretation is that baseline vectors (including those with largest lengths) are foreshortened towards the horizon along their orientation. Thus, they become sensitive to larger angular scales that match the inverse of their foreshortened baseline lengths.

@@

@@

The delay power spectra for these antenna shapes are shown in Figure~\ref{fig:antennas-delay-spectra}. The occupancy of the power patterns on the sky is clearly correlated with that in the delay spectra. The following visual correlations are also apparent --- the strength of the primary lobe centered on the pointing center is correlated with the delay power spectrum centered on $\tau=0$; and, the overall rate of decrease in the power sensitivity away from the pointing center is correlated with the rate of drop in power away from $\tau=0$.

Finally, a distinct feature common to all these aperture shapes is that the foreground contamination near the horizon delay limits even on wide antenna spacings is significant ($\gtrsim 10^5$~K$^2$(Mpc/$h$)$^3$). We have argued this arises due to wide--field effects.

The prevalence of this feature across different antenna shapes demonstrates it is generic to all wide--field measurements, especially in EoR experiments. The amplitude of this effect, however, can be controlled via choice of antenna shape and through weighting of aperture illumination. A dish--shaped antenna appears to hold a significant advantage over a dipole or a phased array from the viewpoint of foreground contamination.

Typically, the sensitivity of antennas to the primary field of view is high compared to the rest of regions on the sky. Combined with the wide--field effects seen earlier, it leads to a ``pitchfork''--shaped signature inside the {\it foreground wedge}, as exemplified in the case of a dish. Although the exact appearance of this signature depends on antenna the power pattern, we use the term ``pitchfork'', hereafter, to broadly refer to the combination of foreground power in the primary field of view and the enhancement of foreground power near the horizon limits due to the nature of wide--field measurements.

>>>>

Equation~\ref{eqn:obsvis} can be {\bf equivalently expressed} as:
\begin{align}\label{eqn:vis}
  V_{uf} &\equiv V_{uf}(\boldsymbol{u},f) \nonumber\\
  &= \iint\limits_\textrm{sky} A(\hat{\boldsymbol{s}},f)\,I(\hat{\boldsymbol{s}},f)\,W_f\,e^{-i2\pi\boldsymbol{u}\cdot\hat{\boldsymbol{s}}}\,\dif\Omega,
\end{align}
where, $\hat{\boldsymbol{s}}$ is measured with reference to a location on the sky referred to as the {\it phase center}, and $\boldsymbol{u}\equiv (u,v,w)$ denotes the spatial frequency vector. $w$ is aligned parallel to the direction of the phase center, while $u$ and $v$ lie on the transverse plane perpendicular to it. For measurements that are constrained to lie on this plane, we can choose $w=0$ without loss of generality and $\boldsymbol{u}$ effectively reduces to $\boldsymbol{u}\equiv (u,v)$, a two--dimensional vector. Then, $\boldsymbol{u}$ is directly related to the transverse spatial wavenumber mode as: 
\begin{align}\label{eqn:k-perp}
  \boldsymbol{k}_\perp &\equiv \frac{2\pi\boldsymbol{u}}{D}, 
\end{align}
where, $D\equiv D(z)$ is the transverse comoving distance at redshift $z$. % Hence, $V_{uf}=V_{uf}(\boldsymbol{b}/\lambda,f)=V_{bf}(\lambda\boldsymbol{u},f)$. It is important to note that for a given antenna pair, $\boldsymbol{b}$ is independent of frequency while $\boldsymbol{u}$ is not.

@@

Without loss of generality, the phase center can be assumed to be the zenith relative to the local tangent plane. Then $\boldsymbol{u}$ lies on this plane for measurements constrained to be on it. If the array of antennas are also coplanar lying on the local tangent plane, then $\boldsymbol{u}=\boldsymbol{b}/\lambda$. Under such circumstances, equations~\ref{eqn:obsvis} and \ref{eqn:delay} closely resemble equations~\ref{eqn:vis} and \ref{eqn:los-transform} respectively. However, they are not quite identical to each other. It is because $\boldsymbol{b}$ is independent of frequency while $\boldsymbol{u}$ is not. \citet{par12b} and \citet{liu14a} have discussed the mathematical correspondence between the two.

@@

{\bf We give a brief overview of some parameters of Fourier space which is generic to all experiments that use a similar approach}. Figure~\ref{fig:fourier-space} illustrates the Fourier space in which the delay (and power) spectra of H{\sc i} from the EoR are presented in this paper. $|\boldsymbol{b}|$ and $k_\perp$, denoting spatial scales in the transverse direction {\bf (tangent plane to the celestial sphere)}, form the $x$--axis. $\tau\approx\eta\propto k_\parallel$, denoting spatial scales along line--of--sight form the $y$--axis. Foreground emission maps to a wedge--shaped region in Fourier space, hereafter referred to as the {\it foreground wedge} \citep{dat10}, whose boundaries are determined by the antenna spacings and the light travel times across them. These boundaries, called horizon delay limits \citep{ved12,par12b}, are shown by solid lines. The spectral transfer function of the instrument convolves the {\it foreground wedge} and stretches it further (unshaded narrow strips bounded by solid and dashed lines) along $\tau$--axis \citep{par12b,thy13}. The width of this narrow strip is inversely proportional to the operating bandwidth. The region of Fourier space excluding the {\it foreground wedge} and the narrow strips is the so--called {\it EoR window}, shown in light and medium shades. {\bf In the context of EoR studies in Fourier space, the H{\sc i} power spectrum from the EoR is expected to decrease rapidly with $|\boldsymbol{k}|$. Hence, the brightest EoR signal will be observed on the shortest baselines and smallest delays. Thus the region of interest for EoR studies is just beyond the horizon delay limits (dashed lines) on short baselines, marked as the region of {\it maximal EoR sensitivity}}.

{\bf In the specific case of the MWA which has a passband constructed using coarse channels, there are period grating responses resulting in repetitions of the {\it foreground wedge} at multiples of 0.78~$\mu$s. Thus with reference to the MWA, the {\it EoR window} is assumed to lie outside the dashed lines but inside the first grating response ($|\tau| \lesssim 0.78\,\mu$s, dot--dashed lines) and is shown in medium shade. The MWA EoR collaboration actively uses this as a test region in Fourier space for diagnostics on data quality}.

\begin{figure}[htb]
\centering
% \includegraphics[width=\linewidth]{figures/v1_0/fourier_space_185.0_MHz_30.7_MHz.eps}
% \includegraphics[width=\linewidth]{f1.eps}
% \includegraphics[width=\linewidth]{fig2.eps}
\includegraphics[width=\linewidth]{figure2.eps}
\caption{Fourier space in which delay (and power) spectra of EoR H{\sc i} signals are presented. The $x$--axis is denoted by $|\boldsymbol{b}|$ (antenna spacing) or $k_\perp$ (transverse wavenumber). The $y$--axis denoted by $\tau$ (delay) or $k_\parallel$ (line--of--sight wavenumber). Here, $k_\perp$ and $k_\parallel$ are obtained for a frequency of 185~MHz. The dark shaded region is referred to as the {\it foreground wedge} where smooth spectrum foregrounds reside. Its boundaries (solid lines), given by light travel time for corresponding antenna spacings, are referred to as horizon delay limits. Regions excluding the {\it wedge} and narrow extensions of the {\it wedge} (white unshaded strips) caused by convolution with the {\bf instrument's spectral transfer function} are expected to be relatively free of foreground contamination {\bf and are generally referred to as the {\it EoR window}. There are undesirable grating responses (dot--dashed lines) specific to the MWA. Hence, we conservatively identify a restricted region of high EoR sensitivity (medium shade) and refer to it as the {\it MWA EoR window}}. \label{fig:fourier-space}}
\end{figure}

In our present study, we use a latitude of -26.701\arcdeg and an antenna layout identical to that of the MWA \citep{bea12}. The array is arranged as a centrally condensed core of $\sim$~300~m --- there are many spacings in the range 5--50~m --- and a radial density that falls off as the inverse of the radius, with the longest baselines at 3~km. Here we focus on antenna spacings $|\boldsymbol{b}| \le 200$~m (spatial scales relevant to reionization). {\bf Their deviation from coplanarity is negligible}. For geometrical intuition, we restrict the orientation ($\theta_b$, measured anti-clockwise from East) of all baselines to lie in the range $-67.5\arcdeg \leq \theta_b < 112.5\arcdeg$. Baselines oriented in the other half--plane measure conjugate visibilities with delays of equal magnitude but of opposite sign and hence are ignored in our analysis. {\bf Note that visibilities from different baselines have not been averaged together in this analysis}.

@@

\begin{figure}[htb]
\centering
\includegraphics[width=\linewidth]{figure3.eps}
\caption{Wide--field effects on delay power spectra produced with a uniform sky brightness distribution measured by antenna pairs with a uniform power pattern across the visible hemisphere. The non--zero response of the interferometer array to a uniform brightness distribution and the prominent {\it edge brightening} close to the horizon delay limits are wide--field effects. These are prevalent on all antenna spacings and are generic to all instruments used in wide--field measurements. \label{fig:usm}}
\end{figure}

We interpret this is due to equal--sized delay bins subtending larger solid angles near the horizon thereby containing larger integrated emission. An alternate, but equivalent, interpretation is that baseline vectors (including those with largest lengths) are foreshortened towards the horizon along their orientation. Thus, they become sensitive to larger angular scales that match the inverse of their foreshortened baseline lengths along these directions.

@@

The model described above is primarily a model of the diffuse foreground sky. While it contains faint compact emission blended in with the diffuse emission, {\bf bright point sources have been removed \citep{deo08}}. In order to supplement it with missing bright compact emission, we use classical radio source confusion estimates to determine the nominal flux density threshold and include point sources brighter than this threshold. It may be noted that slightly different criteria are in common use in radio astronomy to estimate radio source confusion \citep[see Appendix of][and references therein]{thy13}. For an angular resolution of 0.85\arcdeg, using a conservative `$S_\textrm{c}=5\sigma_\textrm{c}$' criterion, we determine the flux density threshold to be $\approx 10$~Jy. Other liberal criteria that yield a lower threshold carry a greater risk of double--counting point sources which might be already blended in with the diffuse sky model.

@@

The delay power spectra without thermal noise component for these antenna shapes are shown in Figure~\ref{fig:antennas-delay-spectra}. The occupancy of the power patterns on the sky is clearly correlated with that in the delay spectra. The following visual correlations are also apparent --- the strength of the primary lobe centered on the pointing center is correlated with the delay power spectrum centered on $\tau=0$; and, the overall rate of decrease in the power sensitivity away from the pointing center is correlated with the rate of drop in power away from $\tau=0$.

The levels of foreground contamination in the {\it EoR window} varies substantially across the different antenna shapes: $\sim 10^4\,$K$^2\,$(Mpc/$h$)$^3$, $\lesssim 10^2\,$K$^2\,$(Mpc/$h$)$^3$, and $<1\,$K$^2\,$(Mpc/$h$)$^3$ while using a dipole, phased array, and a dish, respectively. The severity of foreground contamination inside the {\it foreground wedge} both in strength and occupancy also evidently decreases as the antenna element is changed from a dipole to a phased array to a dish. For instance, notice that the foreground contamination in $k$--modes between $k_\parallel=0$ and the horizon limits decreases from $\sim 10^5$~K$^2\,$(Mpc/$h$)$^3$ in a phased array to $\sim 10$~K$^2\,$(Mpc/$h$)$^3$ in a dish. As a consequence, $k$--modes in the {\it foreground wedge} that may be deemed too contaminated for EoR studies in the case of a dipole or a phased array can potentially become accessible when using a dish.

Finally, a distinct feature common to all these aperture shapes is that the foreground contamination near the horizon delay limits even on wide antenna spacings is significant ($\gtrsim 10^5$~K$^2$(Mpc/$h$)$^3$). We have argued this arises due to wide--field effects. The prevalence of this feature across different antenna shapes demonstrates it is generic to all wide--field measurements, especially in EoR experiments. The amplitude of this effect, however, can be controlled via choice of antenna shape and through weighting of aperture illumination. A dish--shaped antenna appears to hold a significant advantage over a dipole or a phased array from the viewpoint of foreground contamination.

Typically, the sensitivity of antennas to the primary field of view is high compared to the rest of regions on the sky. Combined with the wide--field effects seen earlier, it leads to a ``pitchfork''--shaped signature inside the {\it foreground wedge}, as exemplified in the case of a dish. Although the exact appearance of this signature depends on antenna the power pattern, we use the term {\it pitchfork} hereafter, to broadly refer to the combination of foreground power in the primary field of view and the enhancement of foreground power near the horizon limits due to the nature of wide--field measurements.
